---
title: "CS 7641 UL Report — R Sanity Check"
subtitle: "Python clustering and DR results cross-checked against native R (kmeans/Mclust/prcomp)"
author: "Timothy Leung"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: true
    keep_tex: false
    latex_engine: xelatex
header-includes:
  - \usepackage{xcolor}
  - \usepackage{geometry}
  - \usepackage{titling}
  - \usepackage{titlesec}
  - \usepackage{fancyhdr}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \usepackage{hyperref}
  - \definecolor{denim}{HTML}{2C6FBB}
  - \definecolor{denimdark}{HTML}{1B3A5C}
  - \definecolor{denimmid}{HTML}{3A6B8C}
  - \definecolor{denimbright}{HTML}{5B9BD5}
  - \definecolor{amber}{HTML}{D4A017}
  - \definecolor{plum}{HTML}{7B2D8E}
  - \definecolor{sage}{HTML}{6BB38A}
  - \definecolor{sienna}{HTML}{B8552E}
  - \hypersetup{linkcolor=denimbright, urlcolor=denim, citecolor=sage, filecolor=plum}
  - \geometry{margin=0.7in}
  - \titleformat{\section}{\color{denimdark}\Large\bfseries}{\thesection}{0.6em}{}
  - \titleformat{\subsection}{\color{denimmid}\large\bfseries}{\thesubsection}{0.6em}{}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyhead[L]{\color{denimmid}\small CS 7641 UL Report — R Sanity Check}
  - \fancyhead[R]{\color{denimmid}\small Timothy Leung}
  - \fancyfoot[C]{\color{denimmid}\small \thepage}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.5,
  fig.height = 3.6,
  dpi        = 200,
  cache      = FALSE
)

denim_palette <- list(
  denim      = "#2C6FBB",
  denim_dark = "#1B3A5C",
  denim_mid  = "#3A6B8C",
  amber      = "#D4A017",
  plum       = "#7B2D8E",
  sage       = "#6BB38A",
  sienna     = "#B8552E",
  teal       = "#4C9E82"
)

suppressPackageStartupMessages({
  library(yaml)
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(tibble)
  library(ggplot2)
  library(scales)
  library(knitr)
  have_kableExtra <- requireNamespace("kableExtra", quietly = TRUE)
})

# ggplot2 theme — white bg, whitegrid
denim_theme <- function(base_size = 9) {
  theme_minimal(base_size = base_size, base_family = "sans") +
  theme(
    plot.background   = element_rect(fill = "white", colour = NA),
    panel.background  = element_rect(fill = "white", colour = NA),
    panel.grid.major  = element_line(colour = "#d0d8e8", linewidth = 0.4),
    panel.grid.minor  = element_blank(),
    axis.line         = element_line(colour = denim_palette$denim_dark, linewidth = 0.4),
    axis.text         = element_text(colour = denim_palette$denim_dark),
    axis.title        = element_text(colour = denim_palette$denim_dark, face = "bold"),
    plot.title        = element_text(colour = denim_palette$denim_dark, face = "bold"),
    plot.subtitle     = element_text(colour = denim_palette$denim_mid),
    legend.position   = "right"
  )
}

# Locate project root
project_root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
results_dir  <- file.path(project_root, "results")
logs_dir     <- file.path(results_dir,  "logs")
config_path  <- file.path(project_root, "configs", "config.yaml")
cfg          <- yaml::read_yaml(config_path)
seed         <- as.integer(cfg$experiment$seed)   # 7641
set.seed(seed)
```

# Purpose

This document provides an independent R-side audit of the Python UL pipeline.
It cross-checks:

1. **K-Means internal metrics** — R \texttt{kmeans} vs.\ Python sklearn \texttt{KMeans}.
2. **PCA explained-variance curve** — R \texttt{prcomp} vs.\ Python sklearn \texttt{PCA}.
3. **Clustering log audit** — verifies all Part~1/3 JSON logs are internally consistent.
4. **NN after DR summary** — confirms reported Macro-F1 values are plausible.

> If R \texttt{kmeans} (k=7) silhouette agrees with Python within 0.02, the
> feature matrix and preprocessing are confirmed consistent across implementations.

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# Data — Mirror Python Split

```{r load-data}
cov_csv <- file.path(project_root, "data", "covtype_25k.csv.gz")
df <- if (file.exists(cov_csv)) {
  readr::read_csv(cov_csv, show_col_types = FALSE)
} else {
  NULL
}

if (!is.null(df)) {
  df$Cover_Type <- factor(df$Cover_Type, levels = 1:7)
  cont_cols <- 1:10
  # Scale continuous cols only (mirrors Python)
  mu <- colMeans(df[, cont_cols])
  sg <- apply(df[, cont_cols], 2, sd); sg[sg == 0] <- 1
  df_scaled <- df
  for (j in cont_cols) {
    df_scaled[, j] <- (df_scaled[, j] - mu[j]) / sg[j]
  }
  X_full <- as.matrix(df_scaled[, 1:54])
  y_full <- as.integer(df_scaled$Cover_Type)
}
```

```{r data-summary, results = "asis"}
if (!is.null(df)) {
  tbl <- tibble(
    Split = "Full working sample",
    n = nrow(df),
    n_features = ncol(df) - 1L,
    n_classes = n_distinct(df$Cover_Type)
  )
  k <- knitr::kable(tbl, booktabs = TRUE,
                    caption = "Dataset dimensions (should match Python: 20k × 54).")
  if (have_kableExtra) print(kableExtra::kable_styling(k, latex_options = "HOLD_position", font_size = 9))
  else print(k)
}
```

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# K-Means Sanity Check

```{r helpers}
macro_f1 <- function(truth, pred, lvls = levels(truth)) {
  pred  <- factor(pred,  levels = lvls)
  truth <- factor(truth, levels = lvls)
  cm    <- table(truth, pred)
  f1s   <- sapply(lvls, function(c) {
    tp <- cm[c, c]; fp <- sum(cm[, c]) - tp; fn <- sum(cm[c, ]) - tp
    if (tp == 0) return(0)
    prec <- tp / (tp + fp); rec <- tp / (tp + fn)
    2 * prec * rec / (prec + rec)
  })
  mean(f1s, na.rm = TRUE)
}

silhouette_approx <- function(X, labels, n_sample = 2000, seed = 0) {
  set.seed(seed)
  n  <- nrow(X)
  ix <- if (n > n_sample) sample(n, n_sample) else seq_len(n)
  Xs <- X[ix, ]; ls <- labels[ix]
  if (length(unique(ls)) < 2) return(NA_real_)
  cluster::silhouette(ls, dist(Xs)) |> summary() |> _$avg.width
}
```

```{r r-kmeans, eval = !is.null(df)}
suppressPackageStartupMessages(library(cluster))
set.seed(seed)
km_r <- kmeans(X_full, centers = 7, nstart = 10, iter.max = 300)
r_sil <- silhouette_approx(X_full, km_r$cluster)
```

```{r py-kmeans-load}
py_km_path <- file.path(logs_dir, "kmeans_final.json")
py_km <- if (file.exists(py_km_path)) jsonlite::fromJSON(py_km_path) else NULL
```

```{r kmeans-comparison, results = "asis"}
rows <- tibble(
  Implementation = character(),
  Silhouette = numeric(), ARI = numeric(), Inertia = numeric()
)
if (!is.null(py_km)) {
  rows <- bind_rows(rows, tibble(
    Implementation = "Python sklearn KMeans (k=7)",
    Silhouette = py_km$silhouette,
    ARI        = py_km$ari,
    Inertia    = py_km$inertia
  ))
}
if (exists("r_sil")) {
  rows <- bind_rows(rows, tibble(
    Implementation = "R stats::kmeans (k=7)",
    Silhouette = r_sil,
    ARI        = NA_real_,
    Inertia    = km_r$tot.withinss
  ))
}
if (nrow(rows) > 0) {
  k <- knitr::kable(rows, digits = 4, booktabs = TRUE,
    caption = paste0("K-Means comparison: Python vs. R. ",
                     "|Delta silhouette| < 0.02 = confirmed."))
  if (have_kableExtra) print(kableExtra::kable_styling(k, latex_options = "HOLD_position", font_size = 9))
  else print(k)
}
```

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# PCA Explained Variance Cross-Check

```{r pca-r, eval = !is.null(df)}
pca_r <- prcomp(X_full, center = FALSE, scale. = FALSE)
evr_r <- (pca_r$sdev^2) / sum(pca_r$sdev^2)
cum_r <- cumsum(evr_r)
n90_r <- which(cum_r >= 0.90)[1]
n95_r <- which(cum_r >= 0.95)[1]
```

```{r pca-py-load}
py_pca_path <- file.path(logs_dir, "pca_diagnostics.json")
py_pca <- if (file.exists(py_pca_path)) jsonlite::fromJSON(py_pca_path) else NULL
```

```{r pca-comparison, results = "asis"}
rows_pca <- tibble(
  Implementation = character(),
  `n for 90% var` = integer(),
  `n for 95% var` = integer(),
  `PC1 EVR` = numeric(),
  `PC2 EVR` = numeric()
)
if (!is.null(py_pca)) {
  cum_py <- unlist(py_pca$cum_var)
  evr_py <- unlist(py_pca$evr)
  rows_pca <- bind_rows(rows_pca, tibble(
    Implementation  = "Python sklearn PCA",
    `n for 90% var` = as.integer(which(cum_py >= 0.90)[1]),
    `n for 95% var` = as.integer(which(cum_py >= 0.95)[1]),
    `PC1 EVR`       = evr_py[1],
    `PC2 EVR`       = evr_py[2]
  ))
}
if (exists("pca_r")) {
  rows_pca <- bind_rows(rows_pca, tibble(
    Implementation  = "R prcomp",
    `n for 90% var` = as.integer(n90_r),
    `n for 95% var` = as.integer(n95_r),
    `PC1 EVR`       = evr_r[1],
    `PC2 EVR`       = evr_r[2]
  ))
}
if (nrow(rows_pca) > 0) {
  k <- knitr::kable(rows_pca, digits = 4, booktabs = TRUE,
    caption = "PCA explained variance: Python vs. R. Values should agree within floating-point precision.")
  if (have_kableExtra) print(kableExtra::kable_styling(k, latex_options = "HOLD_position", font_size = 9))
  else print(k)
}
```

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# Clustering Log Audit

```{r log-audit}
# Load all clustering JSON logs
log_files <- list.files(logs_dir,
  pattern = "^(kmeans|gmm)_(final|after_).*.json$",
  full.names = TRUE)

log_rows <- lapply(log_files, function(fp) {
  d   <- jsonlite::fromJSON(fp)
  nm  <- gsub("\\.json$", "", basename(fp))
  tibble(
    log     = nm,
    sil     = d$silhouette,
    ari     = d$ari,
    nmi     = d$nmi,
    ok_sil  = between(d$silhouette, -1, 1),
    ok_ari  = between(d$ari, -1, 1),
    ok_nmi  = between(d$nmi, 0,  1)
  )
}) %>% bind_rows() %>% arrange(log)
```

```{r audit-table, results = "asis"}
disp_audit <- log_rows %>%
  transmute(
    Log     = log,
    Sil     = round(sil, 4),
    ARI     = round(ari, 4),
    NMI     = round(nmi, 4),
    OK      = ifelse(ok_sil & ok_ari & ok_nmi, "✓", "⚠")
  )
k <- knitr::kable(disp_audit, booktabs = TRUE,
  caption = "Clustering log audit: all metric values must be in valid range.",
  align = "lccccc")
if (have_kableExtra) {
  print(kableExtra::kable_styling(k, latex_options = c("HOLD_position", "scale_down"), font_size = 8))
} else {
  print(k)
}
```

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# NN After DR Summary

```{r nn-audit}
nn_path <- file.path(logs_dir, "nn_after_dr.json")
nn_data <- if (file.exists(nn_path)) jsonlite::fromJSON(nn_path) else NULL
```

```{r nn-table, results = "asis"}
if (!is.null(nn_data)) {
  tags <- c("original", "pca", "ica", "rp")
  nn_rows <- lapply(tags, function(tag) {
    r  <- nn_data[[tag]]
    tm <- r$test_metrics
    tibble(
      Features    = tag,
      `dim`       = r$in_dim,
      `Macro-F1`  = tm$macro_f1,
      `Accuracy`  = tm$accuracy,
      `Bal. Acc.` = tm$balanced_acc,
      `Train (s)` = r$wall_train_s,
      `Epochs`    = r$n_epochs_run
    )
  }) %>% bind_rows()

  k <- knitr::kable(nn_rows, digits = 4, booktabs = TRUE,
    caption = "NN after DR — test metrics and runtime. PCA should lead on Macro-F1.")
  if (have_kableExtra) {
    print(kableExtra::kable_styling(k, latex_options = "HOLD_position", font_size = 9))
  } else {
    print(k)
  }
}
```

```{r nn-plot, fig.cap = "NN Macro-F1 by feature space. Positive delta = improvement over original."}
if (!is.null(nn_data)) {
  tags <- c("original","pca","ica","rp")
  labels <- c("Original","PCA","ICA","RP")
  cols   <- c("#1B3A5C","#2C6FBB","#D4A017","#4C9E82")
  f1s    <- sapply(tags, function(t) nn_data[[t]]$test_metrics$macro_f1)
  orig_f1 <- f1s["original"]
  deltas   <- f1s - orig_f1

  df_plot <- tibble(tag = labels, delta = deltas, f1 = f1s, col = cols)

  ggplot(df_plot, aes(x = reorder(tag, f1), y = f1, fill = tag)) +
    geom_col(width = 0.6, colour = NA) +
    geom_text(aes(label = sprintf("%.4f", f1)), hjust = -0.15, size = 2.8,
              colour = "#1A1A2E") +
    scale_fill_manual(values = setNames(cols, labels), guide = "none") +
    coord_flip() +
    labs(x = NULL, y = "Test Macro-F1",
         title = "NN after DR — Test Macro-F1 by feature space") +
    denim_theme()
}
```

\noindent
\textcolor{denimdark}{\rule{\linewidth}{0.6pt}}

# Provenance

- Config: `r normalizePath(config_path)`
- Seed: `r seed`
- Logs dir: `r normalizePath(logs_dir)`
- Clustering logs audited: `r nrow(log_rows)`
- All clustering logs pass range check: `r all(log_rows$ok_sil & log_rows$ok_ari & log_rows$ok_nmi)`
- R version: `r paste(R.version$major, R.version$minor, sep = ".")`
- Knitted: `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`
