% REPRO_UL.tex — CS7641 UL Reproducibility Sheet
% Timothy Leung (tleung37) | gtID 904150400 | seed 7641
% Style mirrors REPRO_SL.tex and REPRO_OL.tex exactly.

\documentclass[11pt]{article}

% ── packages ──────────────────────────────────────────────────────────────────
\usepackage[margin=0.8in]{geometry}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{fancyhdr}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{multirow}

% ── denim palette ─────────────────────────────────────────────────────────────
\definecolor{denim}{HTML}{2C6FBB}
\definecolor{denimdark}{HTML}{1B3A5C}
\definecolor{denimmid}{HTML}{3A6B8C}
\definecolor{denimbright}{HTML}{5B9BD5}
\definecolor{denimrow}{HTML}{E6EEF5}
\definecolor{amber}{HTML}{D4A017}
\definecolor{plum}{HTML}{7B2D8E}
\definecolor{sage}{HTML}{6BB38A}
\definecolor{sienna}{HTML}{B8552E}
\definecolor{parchment}{HTML}{F4EFE5}
\pagecolor{white}

\hypersetup{colorlinks=true, linkcolor=denimbright,
            citecolor=sage, urlcolor=denim}

\lstset{basicstyle=\ttfamily\small, breaklines=true, frame=single,
        rulecolor=\color{denimbright},
        keywordstyle=\color{denim}, commentstyle=\color{sage},
        backgroundcolor=\color{white}}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\color{denimmid}CS7641 UL Reproducibility --- Timothy Leung}
\fancyhead[R]{\small\color{denimmid}gtID 904150400 \textbar\ tleung37@gatech.edu}
\fancyfoot[C]{\small\color{denimmid}\thepage}
\renewcommand{\headrulewidth}{0.3pt}

\setlist[itemize]{leftmargin=*, nosep, topsep=2pt}
\setlist[enumerate]{leftmargin=*, nosep, topsep=2pt}

% ══════════════════════════════════════════════════════════════════════════════
\begin{document}

{\Large\bfseries\color{denimdark}
CS7641 Unsupervised Learning --- Reproducibility Sheet}\\[4pt]
{\large\color{denimmid}
Timothy Leung \quad tleung37@gatech.edu \quad gtID~904150400}\\[2pt]
{\color{denimmid}\rule{\linewidth}{0.6pt}}

\section*{1.\ Code and Project Links}
\begin{itemize}
  \item \textbf{Overleaf (read-only):}
        \url{https://www.overleaf.com/project/6a123a01f7534325be8dfacf}
  \item \textbf{GitHub (GT Enterprise):}
        \url{https://github.gatech.edu/tleung37/CS7641_ml_report3}
  \item \textbf{Commit hash:} \texttt{a3f2aaae04927569652be33535786df0fc2f0d1c}
  \item \textbf{Seed:} \texttt{7641} (course number; propagated everywhere)
\end{itemize}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{2.\ Covertype Sample Continuity}

The working dataset is \textbf{identical} to the SL and OL Reports:
\begin{itemize}
  \item 20,000-instance stratified sample (\texttt{StratifiedShuffleSplit},
        seed~7641) from UCI Forest Cover Type.
  \item Same 60/20/20 train/val/test split (12,000/4,000/4,000 instances).
  \item Preprocessing: \texttt{StandardScaler} on continuous columns 0--9
        (elevation, aspect, slope, hydrology, hillshade, fire distance)
        fit on \textbf{training split only} for downstream NN evaluation.
        For exploratory clustering/DR: fit on full 20k sample.
        Binary columns 10--53 (wilderness, soil) left as 0/1.
  \item \texttt{covtype\_25k.csv.gz} in \texttt{data/} is the same file
        generated by \texttt{report\_1\_sl/src/data\_loader.py} and reused
        in \texttt{report\_2\_ol/} and \texttt{report\_3\_ul/}.
\end{itemize}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{3.\ Environment Setup}

\begin{lstlisting}[language=bash]
# Python environment
python -m venv .venv && source .venv/bin/activate
pip install -r requirements.txt
# Key versions: torch==2.3.1  scikit-learn==1.5.0
#   numpy==1.26.4  pandas==2.2.2  matplotlib==3.9.0
#   seaborn==0.13.2  scipy==1.13.0  pyyaml==6.0
#   umap-learn==0.5.12

# R environment (for rsanity/ audit)
Rscript -e 'install.packages(c("bookdown","yaml","jsonlite","dplyr","ggplot2",
            "knitr","kableExtra","readr","tibble","tidyr","scales","cluster"))'
\end{lstlisting}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{3.2\ Commands to Reproduce Every Figure and Table}

\begin{lstlisting}[language=bash]
# (a) Run the full pipeline end-to-end.
# Reads configs/config.yaml, writes results/{logs,tables,figures}/
python run_pipeline.py

# (b) Compile the UL report PDF (twice for cross-references).
cd report
pdflatex -interaction=nonstopmode UL_Report.tex
pdflatex -interaction=nonstopmode UL_Report.tex
cd ..

# (c) Rerun the R sanity check (independent audit of Python results).
Rscript -e 'rmarkdown::render("rsanity/sanity_check_UL.Rmd", output_dir="rsanity")'

# (d) Compile this reproducibility sheet.
cd repro
pdflatex -interaction=nonstopmode REPRO_UL.tex
\end{lstlisting}

\subsection*{3.3\ Data Paths}

\texttt{sklearn.datasets.fetch\_covtype()} is called by
\texttt{src/data\_loader.py} on first run if \texttt{data/covtype\_25k.csv.gz}
is absent; the dataset ($\sim$75\,MB) caches to \texttt{.cache/covtype/}.
No manual download is required.

\subsection*{3.4\ Random Seeds}

Primary seed: \texttt{seed~=~7641} (course number; same as SL/OL Reports),
set in \texttt{configs/config.yaml} and propagated to
\texttt{numpy.random.seed}, \texttt{random.seed},
\texttt{torch.manual\_seed}, and CuDNN determinism flags.
RP seed sweep: \texttt{[7641, 42, 123, 7, 2024]}.
ICA seed sensitivity sweep: \texttt{[7641, 42, 123, 7, 2024]}.

\subsection*{3.5\ Output Landing Locations}
\begin{itemize}
  \item \texttt{results/logs/eda\_summary.json} --- dataset dimensions and class counts.
  \item \texttt{results/logs/kmeans\_sweep.json} --- K-Means sweep metrics.
  \item \texttt{results/logs/kmeans\_final.json} --- K-Means final (k=7) metrics + contingency.
  \item \texttt{results/logs/gmm\_sweep.json} --- GMM sweep (AIC, BIC, ARI).
  \item \texttt{results/logs/gmm\_final.json} --- GMM final (n=7) metrics + contingency.
  \item \texttt{results/logs/pca\_diagnostics.json} --- EVR, cumulative variance, RMSE.
  \item \texttt{results/logs/ica\_diagnostics.json} --- kurtosis, seed sensitivity.
  \item \texttt{results/logs/rp\_diagnostics.json} --- distance-preservation across 5 seeds.
  \item \texttt{results/logs/umap\_diagnostics.json} --- UMAP hyperparameters + wall time.
  \item \texttt{results/logs/kmeans\_after\_\{pca,ica,rp\}.json} --- Part~3 KMeans.
  \item \texttt{results/logs/gmm\_after\_\{pca,ica,rp\}.json} --- Part~3 GMM.
  \item \texttt{results/logs/nn\_after\_dr.json} --- Part~4 NN metrics for all 4 spaces.
  \item \texttt{results/tables/clustering\_summary.csv} --- compact clustering comparison.
  \item \texttt{results/tables/nn\_dr\_summary.csv} --- compact NN comparison.
  \item \texttt{results/figures/*.pdf} --- all 10 figures (white bg, denim palette).
  \item \texttt{rsanity/sanity\_check\_UL.pdf} --- R audit output.
\end{itemize}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{4.\ Clustering Disclosures (Part~1)}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
K-Means k range   & 2, 3, 4, 5, 6, 7, 8, 10, 12 \\
K-Means final k   & 7 (matches label count) \\
K-Means n\_init   & 20 \\
K-Means max\_iter & 500 \\
GMM n range       & 2, 3, 4, 5, 6, 7, 8, 10, 12 \\
GMM final n       & 7 \\
GMM covariance    & full \\
GMM n\_init       & 5 \\
GMM max\_iter     & 200 \\
Silhouette sample & 3,000 (random subsample for speed) \\
Seed              & 7641 \\
\bottomrule
\end{tabular}
\end{center}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{5.\ DR Disclosures (Part~2)}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Method} & \textbf{Parameter} & \textbf{Value} \\
\midrule
PCA & n\_components (full) & 54 (all) \\
PCA & n\_components (final) & 30 (99.5\% cumvar) \\
ICA & n\_components         & 20 \\
ICA & max\_iter / tol        & 500 / 1e-4 \\
ICA & whiten                 & unit-variance \\
ICA & seed sweep             & [7641, 42, 123, 7, 2024] \\
RP  & n\_components          & 30 \\
RP  & type                   & GaussianRandomProjection \\
RP  & seed sweep             & [7641, 42, 123, 7, 2024] \\
RP  & dist-preservation sample & 500 pairs \\
UMAP (extra) & n\_components  & 2 (visualization only) \\
UMAP (extra) & n\_neighbors    & 30 \\
UMAP (extra) & min\_dist       & 0.1 \\
\bottomrule
\end{tabular}
\end{center}

UMAP was used for visualization only.
It was \textbf{not} used for downstream NN features because UMAP does not
support standard transductive train/test transform without fitting on
the test set.

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{6.\ NN after DR Disclosures (Part~4)}

\textbf{Best justified recipe from SL and OL Reports:}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Architecture      & [256, 128] hidden, ReLU, BatchNorm \\
Optimizer         & Adam no-bias-correction (OL Part2 best) \\
Learning rate     & $10^{-3}$ \\
Weight decay (L2) & $10^{-4}$ (OL Part3 best single regularizer) \\
Batch size        & 256 \\
Epochs            & 40 (max) \\
Early stopping    & patience = 10, monitor val loss \\
Seed              & 7641 \\
DR leakage control & fit on train split only, apply to val/test \\
\bottomrule
\end{tabular}
\end{center}

Input dimension adjusted per DR method:
original~54, PCA~30, ICA~20, RP~30.
All other parameters held constant for fair comparison.

Adam without bias-correction update rule:
\begin{align*}
m_t &= \beta_1 m_{t-1} + (1-\beta_1)g_t, \\
v_t &= \beta_2 v_{t-1} + (1-\beta_2)(g_t)^{2}, \\
\theta_t &= \theta_{t-1} - \alpha\, m_t / (\sqrt{v_t}+\varepsilon).
\end{align*}
Custom implementation in \texttt{src/nn\_after\_dr.py::AdamNoBias}.

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{7.\ R Sanity Check --- Method and Interpretation}

The \texttt{rsanity/sanity\_check\_UL.Rmd} performs three audits:

\begin{enumerate}
  \item \textbf{K-Means agreement.} Runs R \texttt{stats::kmeans} (k=7,
        nstart=10) on the same scaled 20k sample and compares silhouette
        to Python sklearn. Observed: R~$=$~0.1322 vs.\ Python~$=$~0.1306
        ($|\Delta|{=}0.0016 < 0.02$ --- \textbf{confirmed}).
  \item \textbf{PCA agreement.} R \texttt{prcomp} vs.\ Python PCA:
        $n_{90\%}{=}10$, $n_{95\%}{=}14$, PC1-EVR~$=$~0.226 --- exact match.
  \item \textbf{Log audit.} Loads all 10 clustering JSON logs
        (\texttt{kmeans/gmm\_\{final,after\_pca,after\_ica,after\_rp\}.json})
        and verifies silhouette $\in [-1,1]$, ARI $\in [-1,1]$,
        NMI $\in [0,1]$. \textbf{All 10 logs pass.}
\end{enumerate}

{\color{denimmid}\rule{\linewidth}{0.4pt}}

\section*{8.\ Extended Clustering Table}

\begin{center}
\small
\begin{tabular}{llrrrrr}
\toprule
\textbf{Space} & \textbf{Algo} & \textbf{Sil} & \textbf{DB} &
\textbf{CH} & \textbf{ARI} & \textbf{NMI} \\
\midrule
Original & K-Means & 0.1306 & 1.792 & 2683 & 0.023 & 0.067 \\
Original & GMM     & $-$0.007 & 4.087 & 513 & 0.052 & 0.131 \\
PCA-30   & K-Means & 0.1326 & 1.782 & --- & 0.022 & 0.066 \\
PCA-30   & GMM     & $-$0.001 & 3.501 & --- & 0.117 & 0.186 \\
ICA-20   & K-Means & 0.1416 & 1.701 & --- & $-$0.005 & 0.105 \\
ICA-20   & GMM     & 0.1304 & 1.780 & --- & 0.027 & 0.143 \\
RP-30    & K-Means & 0.1478 & 1.650 & --- & 0.027 & 0.080 \\
RP-30    & GMM     & 0.013  & 3.402 & --- & 0.125 & 0.170 \\
\bottomrule
\end{tabular}
\end{center}

\section*{9.\ Extended NN After DR Table}

\begin{center}
\small
\begin{tabular}{lrrrrr}
\toprule
\textbf{Features} & \textbf{dim} & \textbf{Macro-F1} &
\textbf{Accuracy} & \textbf{Bal.~Acc.} & \textbf{Train~(s)} \\
\midrule
Original & 54 & 0.6803 & 0.7940 & 0.6508 & 5.6 \\
PCA-30   & 30 & 0.6951 & 0.8000 & 0.6642 & 7.0 \\
ICA-20   & 20 & 0.6785 & 0.7875 & 0.6359 & 7.2 \\
RP-30    & 30 & 0.6670 & 0.7883 & 0.6339 & 6.9 \\
\bottomrule
\end{tabular}
\end{center}

\section*{10.\ Hardware and Software Versions}
\begin{itemize}[nosep]
  \item OS: Ubuntu / Debian Linux (x86\_64)
  \item CPU: 2-core x86\_64; no GPU used for submission numbers
  \item Python: 3.12.x
  \item Key packages: \texttt{torch~2.3.1}, \texttt{scikit-learn~1.5.0},
        \texttt{numpy~1.26.4}, \texttt{pandas~2.2.2},
        \texttt{matplotlib~3.9.0}, \texttt{seaborn~0.13.2},
        \texttt{scipy~1.13.0}, \texttt{umap-learn~0.5.12}
  \item R: 4.5.0; \texttt{cluster~2.1}, \texttt{bookdown~0.40},
        \texttt{jsonlite~1.8}, \texttt{ggplot2~3.5}, \texttt{kableExtra~1.4}
  \item \LaTeX: TeX Live 2025 / Debian with IEEEtran conference class
\end{itemize}

\end{document}
